A portable, annotation-based, visual stepper for Common Lisp

### 2020-04-07 paper/paper.tex
Zenodo ID thing
Published

\documentclass[sigconf]{acmart}

@@

\author{João Távora} \affiliation{Ravenpack}
\email{joaotavora@gmail.com}
\acmConference{2020 European Lisp Symposium}{April 27th - April 28th 2020}{Zurich, Switzerland}

\setcopyright{rightsretained} % if you give the rights to ACM
% \acmDOI{...} % DOI - Insert your DOI below...
% \acmISBN{...} % ISBN - Insert your conference/workshop's ISBN below...
\acmYear{2020} % Insert Publication year
\copyrightyear{2020} % Insert Copyright year (typically the same as above)
\acmConference[ELS 2020]{2020 European Lisp Symposium}{April 27th - April 28th 2020}{Zurich, Switzerland}

@@

The system as been described as ``portable'' or ``near-portable''.
Indeed, if a completely new conforming implementation of Common Lisp
were to spring into being, support for our stepper system would have
to deal with three potential sources of non-portability: (1) support
of the \texttt{AGNOSTIC-LIZARD} code-walking program, (2) differing
representations of constituent characters in readtables (for the
source-tracking reader described in section \ref{st-reader}), and (3)
support for ``gray streams''.  We defer the discussion of (1) to
\cite{agnostic-lizard}, noting that that system is built with
portability as its foremost requisite.  The ``shielding'' it offers is
useful e.g. in dealing with SBCL's implementation of
\texttt{CL:DEFMETHOD}, which itself produces a complicated
transformation of its body.  In (2), we note that the adjustments to
the read-table were needed only for SBCL's implementation, who doesn't
use a the macro-character function for constituents.  If the
hypothetical new implementation \emph{also} did so, it is plausible
that the current code would also support it.  Otherwise, if it were to
use a function (as the remaining implementations), it would probably
require no extra work as well.  Lastly, for (3), we think it
reasonable to expect support for ``gray streams'' in new
implementations, since it is a widely adopted extension, and required
by SLY/Emacs to begin with.

@@

The SLY/Emacs stepper described in this essay is an effective example
of a portable, visual stepping facility for the Common Lisp ecosystem.
To the best of our knowledge, it is indeed the \emph{only} system
combining these characteristics.  As such, there is little to compare
it against.  The closest match could well lie outside of Common Lisp,
in the aforementioned \texttt{edebug.el}\cite{edebug} stepper, an
annotation-based approach that is equally well integrated with the
source code editor.  That stepper has a better interface, but requires
custom declarations for stepping into macro expansions, something the
SLY/Emacs stepper handles automatically.  Its current limitations
notwithstanding, we believe that the underpinnings of the SLY/Emacs
stepper -- \emph{stickers}, a simple source-tracking reader, and
mnesic macroexpansion -- are solid.  We envision enhancements to its
interface, perhaps by incorporating ideas of non-portable visual
steppers, or steppers for other programming languages.

>>>>

\documentclass[format=sigconf]{acmart}
\bibliographystyle{plainnat}

@@

\author{João Távora} \affiliation{Ravenpack}
\email{joaotavora@gmail.com}
\acmConference[ELS’20]{the 13th European Lisp Symposium}{April 27--28 2020}{%
Zürich, Switzerland}
\acmDOI{10.5281/zenodo.3742759}

\setcopyright{rightsretained} % if you give the rights to ACM
% \acmDOI{...} % DOI - Insert your DOI below...
% \acmISBN{...} % ISBN - Insert your conference/workshop's ISBN below...
\acmYear{2020} % Insert Publication year
\copyrightyear{2020} % Insert Copyright year (typically the same as above)

@@

The system as been described as ``portable'' or ``near-portable''.
Indeed, if a completely new conforming implementation of Common Lisp
were to spring into being, support for our stepper system would have
to deal with three potential sources of non-portability: (1) support
of the \texttt{AGNOSTIC-LIZARD} code-walking program, (2) differing
representations of constituent characters in readtables (for the
source-tracking reader described in section \ref{st-reader}), and (3)
support for ``gray streams''.  We defer the discussion of (1) to
\cite{agnostic-lizard}, noting that that system is built with
portability as its foremost requisite.  The ``shielding'' it offers is
useful e.g. in dealing with SBCL's implementation of
\texttt{CL:DEFMETHOD}, which itself produces a complicated
transformation of its body.  In (2), we note that the adjustments to
the read-table were needed only for SBCL's implementation, which
doesn't use a macro-character function for constituents.  If the
hypothetical new implementation \emph{also} did so, it is plausible
that the current code would support it.  Otherwise, it would behave as
the remaining implementations, also requiring no extra work.  Lastly,
for (3), we think it reasonable to expect support for ``gray streams''
in new implementations, since it is a widely adopted extension, and
required by SLY/Emacs to begin with.

@@

The SLY/Emacs stepper described in this essay is an effective example
of a portable, visual stepping facility for the Common Lisp ecosystem.
To the best of our knowledge, it is indeed the \emph{only} system
combining these characteristics.  As such, there is little to compare
it against.  The closest match could well lie outside of Common Lisp,
in the aforementioned \texttt{edebug.el}\cite{edebug} stepper, an
annotation-based approach that is equally well integrated with the
source code editor.  That stepper has a more developed interface, but
requires custom declarations for stepping into macro expansions,
something the SLY/Emacs stepper handles automatically.  Its current
limitations notwithstanding, we believe that the underpinnings of the
SLY/Emacs stepper -- \emph{stickers}, a simple source-tracking reader,
and mnesic macroexpansion -- are solid.  We envision enhancements to
its interface, perhaps by incorporating ideas of non-portable visual
steppers, or steppers for other programming languages.
